---
title: "Chapter 5: Dimensionality reduction techniques"
author: "Anni Norring"
date: "30 marraskuuta 2017"
output: html_document
---

# Chapter 5: Dimensionality reduction techniques

## RStudio exercise 5: Data wrangling and analysis

## Data wrangling

You can find my [preprocessed data set](https://github.com/anorring/IODS-project/blob/master/human.csv) and the [data wrangling R script](https://github.com/anorring/IODS-project/blob/master/data/create_human.R) from my [GitHub repository.](https://github.com/anorring/IODS-project/tree/master/data)

## Analysis


```{r, include=FALSE}
# Access the needed libraries:
library(dplyr)
library(tidyr)
library(ggplot2)
library(boot)
library(MASS)
library(tidyverse)
library(corrplot)
```

### 1. Data

In this week assignment we are working with data from the the Human Development Index (HDI) and the Gender Inequality Index (GII), which are both compiled by the United Nations Development programme. You can learn more about the data [here on HDI](http://hdr.undp.org/en/composite/HDI) and [here on GII](http://hdr.undp.org/en/composite/GII). 

The dataset we will use was combined and modified in the data wrangling part of this assignment. 

We will start by loading the preprocessed "human" data into R and taking a look at it's general appearance. 

```{r}
# load the data
human <- read.table("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/human2.txt", sep = ",", col.names = TRUE, row.names = TRUE)

# glimpse at the dataset
glimpse(human)
```

